Climate change projected to reduce European life expectancy

### 2018-03-21 MS-cclifeexpec.Rmd
minor edits.

---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: LATEX/svm-latex-ms.tex
title: "Climate change projected to reduce European life expectancy"
thanks: "All data and code that supports these conclusions are available upon request at hauer@uga.edu."
author:
- name: Mathew E. Hauer
  affiliation: University of Georgia
- name: Alexis R. Santos
  affiliation: Pennsylvania State University

abstract: "Climate change related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change. However, life expectancy at birth is more easily communicated to the public, understood by the public, and controls for different geographic, economic, and demographic groups. No studies have comprehensively connected excess mortality due to climate change to anticipated reductions in life expectancy. Without properly situating the potential loss of life within the contexts of life expectancy, we risk misrepresenting cost of climate change on mortality. In this paper, we convert excess mortality estimates into potential reductions of life expectancy at birth and report the cost of climate change on human longevity. We find climate change could become the third largest life expectancy reducer behind heart disease and cancer in some European countries. Thus, the cost of inaction on climate change could approach one year of life in some countries."

keywords: "Climate change, Life expectancy, Mortality, Demography, Excess mortality, Europe, Public health"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
#fontfamily: mathpazo
fontsize: 12pt
spacing: double
bibliography: LATEX/ccmort.bib
biblio-style: nature
fig_caption: true
# use apsr or nature
header-includes:
- \usepackage[all]{nowidow}
- \usepackage{rotating}
---

@@

# Methods
We estimate changes in life expectancy by combining three primary datasets: excess mortality data from Forzieri et al. [@forzieri2017increasing], cause-of-death distributions from the Global Burden of Disease study (GBD) [@GBD;@wang2012age], and life tables from the Human Mortality Database (HMD) [@HMD]. Forzieri et al. [@forzieri2017increasing] focused on the projected excess mortality of heatwaves, coldwaves, wildfires, droughts, river and coastal floods, and windstorms under the business-as-usual greenhouse gas emission scenario. They modelled long-term, small-area demographic changes in European countries that correspond to a middle-of-the-road socioeconomic scenario and based their projected mortality on extrapolations of an exhaustive examination of contemporary disaster databases. Their results represent business-as-usual climate change and human development without incorporating potential adaptation. They found a potential 150,000+ climate change related fatalities per year by the mid 2080s with climate change contributing to 90% of mortality rise as opposed to population changes. These data provide the anticipated fatalities but not the effect these deaths might have on life expectancy.

@@

# Results
We find that climate change could alter life expectancy by -0.22 years (-2.5 months) in the average European country (**Fig. 1**). This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States. Although the average European country could see a change of -0.22 years, several countries are likely to experience considerably greater reductions in life expectancy. Luxembourg could see a change of up to -2.2 years and the medium-variant of climate hazards could cost the average Spaniard 0.98 years of life.

@@

While Forzieri et al's [@forzieri2017increasing] results suggest that the greatest climate change related mortality will unfold along a north-south gradient, we find the greatest reduction in life expectancy unfolds along an east-west gradient (**Fig. 2**). The most westerly European countries tend to have the greatest reductions. Neighboring pairs of countries at similar latitudes could experience vastly different mortality regimes, but neighboring pairs of countries at similar longitudes seem to experience more similar mortality regimes from climate change. Additionally, these results point to the importance of converting excess mortality into life expectancy to properly quantify the effects of climate change mortality.

@@

Reductions such as those should not be taken lightly. Many of the children born today are likely to still be alive by the end of the century and will be in the age groups most threatened by extreme heat - the biggest mortality risk associated with climate change [@keatinge2000heat]. If climate change unfolds as a more aggressive mortality vector, only major cardiovascular diseases and malignant neoplasms would contribute more to life expectancy reductions in numerous European countries [@arias2013united]. This would make climate change one of the most aggressive new mortality vectors to emerge over the last quarter-century, representing a major threat to public health in many parts of Europe.

@@

# Supplemenatary Information
## Data
To create life expectancy differentials, we used three sources of data. The first is the baseline life table data. Baseline life tables comes from the Human Mortality Database (HMD) [@HMD]. These datasets comprise only official vital event registration and are considered the most complete and accurate source of mortality data available. We downloaded the most recent 1x5 (by age and year) life table data for twenty-eight European countries. We used the following data elements from the five-year life table data in the HMD:

@@

## Methods
We derived a number of additional variables to accomplish our analysis. First, we abridged the life table data from the HMD from 100+ to 80+ to conform to the GBD cause-specific mortality schedules.

From the GBD data, we derived variable $t_{x,i}$ as the proportion of deaths, $D$, from each age group $x$ in each country $i$ ($D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_i^{GBD}}$).

@@

$$ _nm_{x,i,BASE} = _nD_{x,i}^{HMD} / _nP_{x,i}^{HMD} $$
$$ _n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + \hat{D_i} \cdot _nt_{x,i} ) / _nP_{x,i}^{HMD} $$

@@

$$_nq_{x,i,BASE} = \frac{m_{x,i,BASE}}{1+(n-_na_{x,i}) \cdot m_{x,i,BASE}}$$
$$_n\hat{q}_{x,i,s} = \frac{\hat{m}_{x,i,s}}{1+(n-_na_{x,i}) \cdot \hat{m}_{x,i,s}}$$

We calculated each additional life table value identically for each scenario $s$ using the standard life table equations:

>>>>

---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: true
    fig_caption: true
    latex_engine: pdflatex
    template: LATEX/svm-latex-ms.tex
title: "Climate change projected to reduce European life expectancy"
thanks: "All data and code that supports these conclusions are available as supplementary materials."
author:
- name: Mathew E. Hauer^1^*
  affiliation: University of Georgia
- name: Alexis R. Santos^2^
  affiliation: Pennsylvania State University

abstract: "Climate change related estimates of excess mortality clearly demonstrate the dramatic impact on public health and human mortality from climate change. However, life expectancy at birth is more easily communicated and understood by the public, and controls for different geographic, economic, and demographic groups. No studies have comprehensively connected excess mortality due to climate change to anticipated reductions in life expectancy. Without properly situating the potential loss of life within the contexts of life expectancy, we risk misrepresenting cost of climate change on mortality. In this paper, we convert excess mortality estimates into potential reductions of life expectancy at birth and report the cost of climate change on human longevity. We find climate change could become the third largest life expectancy reducer behind heart disease and cancer in some European countries. Thus, the cost of inaction on climate change could approach one year of life in some countries."

keywords: "Climate change, Life expectancy, Mortality, Demography, Excess mortality, Europe, Public health"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
#fontfamily: mathpazo
fontsize: 12pt
spacing: double
bibliography: LATEX/ccmort.bib
biblio-style: nature
fig_caption: true
# use apsr or nature
header-includes:
- \usepackage[all]{nowidow}
- \usepackage{rotating}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead{}
- \fancyhead[CE,CO]{Climate change life expectancy}
---

* Corresponding author. hauer@uga.edu. p: 706-542-9369.

^1^ Carl Vinson Institute of Government, University of Georgia. 201 N. Milledge Ave. Athens, GA USA 30602.

^2^ Department of Sociology and Criminology, Pennsylvania State University. 901 Oswald Tower, University Park, PA USA 16802.

@@

# Methods
We estimate changes in life expectancy by combining three primary datasets: excess mortality data from Forzieri et al. [@forzieri2017increasing], cause-of-death distributions from the Global Burden of Disease study (GBD) [@GBD;@wang2012age], and life tables from the Human Mortality Database (HMD) [@HMD]. Forzieri et al. [@forzieri2017increasing] focused on the projected excess mortality of heatwaves, coldwaves, wildfires, droughts, river and coastal floods, and windstorms under the business-as-usual greenhouse gas emission scenario (SRES A1B). They modelled long-term, small-area demographic changes in European countries that correspond to a middle-of-the-road socioeconomic scenario (Shared-Socioeconomic Pathway 2 with medium fertility, medium mortality, medium migration, and the Global Education Trend education scenario) and based their projected mortality on extrapolations of an exhaustive examination of contemporary disaster databases within spatially rectified, high-resolution gridded datasets. While the disaster data are not standardized across country, Forzeri et al. imputed data in incomplete time periods and countries. This imputation could mask spatial variability at the sub-national level, but here we use their country-aggregated results. Their results represent business-as-usual climate change and human development without incorporating potential adaptation. They found a potential 150,000+ climate change related fatalities per year by the mid 2080s with climate change contributing to 90% of mortality rise as opposed to population changes. These data provide the anticipated fatalities but not the effect these deaths might have on life expectancy.

@@

# Results
We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.5 months (`r round(mean(lt_lancet$DIF_LOW),2)` - `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**Fig. 1**). This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States. Although the average European country could see a change of -0.22 years, several countries are likely to experience considerably greater reductions in life expectancy. Luxembourg could see a change of up to -2.2 years and the medium-variant of climate hazards could cost the average Spaniard 0.98 years of life.

@@

While Forzieri et al's [@forzieri2017increasing] results suggest that the greatest climate change related mortality will unfold along a north-south gradient, we find the greatest reduction in life expectancy unfolds along an east-west gradient (**Fig. 2**). The most westerly European countries tend to have the greatest reductions. Neighboring pairs of countries at similar latitudes could experience vastly different mortality regimes, but neighboring pairs of countries at similar longitudes seem to experience more similar mortality regimes from climate change. Emergent research in Atmospheric Rivers suggest Western Europe could be more susceptible to these types of events [@ramos2015daily]. Additionally, these results point to the importance of converting excess mortality into life expectancy to properly quantify the effects of climate change mortality.

@@

Reductions such as those should not be taken lightly. Many of the children born today are likely to still be alive by the end of the century and will be in the age groups most threatened by extreme heat (65+) - the biggest mortality risk associated with climate change [@keatinge2000heat]. If climate change unfolds as a more aggressive mortality vector, only major cardiovascular diseases and malignant neoplasms would contribute more to life expectancy reductions in numerous European countries [@arias2013united]. This would make climate change one of the most aggressive new mortality vectors to emerge over the last quarter-century, representing a major threat to public health in many parts of Europe.

@@

# Supplemenatary Information
## Data
To create life expectancy differentials, we used three sources of data. The first is the baseline life table data. Baseline life tables comes from the Human Mortality Database (HMD) [@HMD]. These datasets comprise only official vital event registration and are considered the most complete and accurate source of mortality data available. We downloaded the most recent 5x1 (by age and year) life table data for twenty-eight European countries. We used the following data elements from the five-year age group life table data in the HMD:

@@

```{r suppfig, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("**Mortality distribution of Heat-related mortality in 28 European countries**. The thick black line is the average\\label{suppfig}")}
GBD_data <- read.csv("data/GBD_data.csv") %>%
  #mutate(perdist = log(val)) %>%
  group_by(countrycode, Age) %>%
  summarise(meanval = mean(val),
            upperval = mean(upper),
            lowerval = mean(lower)) %>%
  group_by(countrycode) %>%
  mutate(countmeanval = sum(meanval),
         countupperval = sum(upperval),
         countlowerval = sum(lowerval),
         perdist = (meanval / countmeanval)) %>%
  rename(CNTRY = countrycode)

ggplot(data=GBD_data, aes(x=Age, y=perdist)) +
  geom_line(aes(color=CNTRY), alpha=1) +
  geom_smooth(span=0.3, se=F, col="black", lwd =2) +
  theme_bw() +
  theme(legend.position = c(0.8,0.25),
        legend.title=element_blank(),
        legend.text=element_text(size=4),
        legend.key.size = unit(1,"line")) +
  guides(col = guide_legend(ncol = 4)) +
  scale_y_log10()+
  labs(y="log(Mortality Percentage Distribution)",
       #title = "Mortality distribution of Heat-related mortality in 28 European countries",
       caption = "Source: Global Burden of Disease")

```

## Methods
We derived a number of additional variables to accomplish our analysis. First, we abridged the life table data from the HMD from ages 100+ to 80+ to conform to the GBD cause-specific mortality schedules.

From the GBD data, we derived variable $t_{x,i}$ as the proportion of deaths, $D$, from each age group $x$ in each country $i$ ($_nt_{x,i}=D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_i^{GBD}}$).

@@

\begin{equation}
_nm_{x,i,BASE} = _nD_{x,i}^{HMD} \,/\, _nP_{x,i}^{HMD}
\end{equation}

\begin{equation}
_n\hat{m}_{x,i,s} = (_nD_{x,i}^{HMD} + (\hat{D_i} \,\cdot\, _nt_{x,i}) ) \,/\, _nP_{x,i}^{HMD}
\end{equation}

@@

\begin{equation}
_nq_{x,i,BASE} = \frac{m_{x,i,BASE}}{1+(n-_na_{x,i}) \cdot m_{x,i,BASE}}
\end{equation}

\begin{equation}
_n\hat{q}_{x,i,s} = \frac{\hat{m}_{x,i,s}}{1+(n-_na_{x,i}) \cdot \hat{m}_{x,i,s}}
\end{equation}

We calculated each additional life table value identically for each scenario $s$ using standard life table equations:
